Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-28 VegOzone.tex
Restructure text

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product \textbf{TODO: reference} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 was one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR).
For comparison, the default values ($A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$) indicating a substantially longer growing season. Due to a lack of quantitative field observation, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland \citep[][Supplement~Fig.~S1, Rovaniemi]{FCR:Korhonen2018}.

@@

Soil texture influences water availability and thus $\mathrm{POD_y}$ calculations in $\mathrm{DO_3SE}$ \citep{ACP:Bueker2012}. The soil at Svanhovd is characterized in an ICP Forest plot survey as gleyic cambisols according to the Food and Agriculture Organization of the United Nations (FAO) soil classification system and as eluviated dystric brunisol in the ICP Forest soil classification system (Volkmar Timmerman, pers. comm.) Within the $\mathrm{DO_3SE}$ soil parameterization, the sandy loam texture describes the upper $60\,\unit{cm}$, where the plant roots are found, of this soil type best.

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are probably smaller than the $5\,\unit{cm}$ from the MM parameterization of silver birch leaves. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances and hence the uptake of \chem{O_3}. In addition, one Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ \textbf{TODO:(Volkmar Timmerman, priv. comm.)}. Canopy height and leaf size are adjusted to local values in the models with bespoke growing season (GS bespoke) in the following.

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for the three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using the generic MM parameterization and our bespoke temperature (\emph{cold}, \emph{subarctic}) and light (PPFD0.8, PPFD1.2) response functions. We also locally adapted leaf size for deciduous trees and the tree height together with the bespoke GS. The results displayed in Fig.~\ref{fig:pody_rel} allow for an assessment of systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions.
For comparison, critical levels for reduction in biomass are shown as horizontal dashed lines \citep[deciduous forest $4\,\unit{\%}$, coniferous forest $2\,\unit{\%}$, and grasslands $10\,\unit{\%}$ biomass reduction,][]{ICP:MappingManual2017,ESPR:Hayes2021}. We identify that an increase in cold tolerance leads to higher ozone uptake in both probed years, while the usage of locally adapted GS causes a reduction. The earlier and longer GS in 2019 leads to increased ozone uptake compared to 2018 despite more episodes of elevated \chem{[O_3]} in 2018. Due to the shape of the light response function, a variation of the extent of stomatal opening at low light intensities leads to an asymmetric response in $\mathrm{POD_1}$ and can cause a higher sensitivity to drought conditions. Similarly, the temperature acclimation to cold or subarctic conditions can lead to an amplification of effects by droughts.

>>>>

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 was one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR).
For comparison, the default values for central Europe ($A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

@@

Soil texture influences water availability and thus $\mathrm{POD_y}$ calculations in $\mathrm{DO_3SE}$ \citep{ACP:Bueker2012}. The soil at Svanhovd is characterized in an ICP Forest plot survey as gleyic cambisols according to the Food and Agriculture Organization of the United Nations (FAO) soil classification system and as eluviated dystric brunisol in the ICP Forest soil classification system (V. Timmerman, personal communication, 2021 May 10) Within the $\mathrm{DO_3SE}$ soil parameterization, the sandy loam texture describes the upper $60\,\unit{cm}$, where the plant roots are found, of this soil type best.

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are probably smaller than the $5\,\unit{cm}$ from the MM parameterization of silver birch leaves. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances and hence the uptake of \chem{O_3}. In addition, one Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ (V. Timmerman, personal communication, 2021 May 10).

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for the three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using the generic MM parameterization and our bespoke temperature (\emph{cold}, \emph{subarctic}) and light (PPFD0.8, PPFD1.2) response functions. Canopy height and leaf size are adjusted to local values in the model simulations with bespoke growing season (GS besp.). The results displayed in Fig.~\ref{fig:pody_rel} allow for an assessment of systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions.
For comparison, critical levels for reduction in biomass are shown as horizontal dashed lines \citep[deciduous forest $4\,\unit{\%}$, coniferous forest $2\,\unit{\%}$, and grasslands $10\,\unit{\%}$ biomass reduction,][]{ICP:MappingManual2017,ESPR:Hayes2021}. We identify that an increase in cold tolerance (MM $\rightarrow$ cold $\rightarrow$ subarctic) leads to higher ozone uptake in both probed years, while the usage of locally adapted GS causes a reduction. The earlier and longer GS in 2019 leads to increased ozone uptake compared to 2018 despite more episodes of elevated \chem{[O_3]} in 2018. Due to the shape of the light response function, a variation of the extent of stomatal opening at low light intensities leads to an asymmetric response in $\mathrm{POD_1}$ and can cause a higher sensitivity to drought conditions. Similarly, the temperature acclimation to cold or subarctic conditions can lead to an amplification of effects by droughts.
